AMORPH: A statistical program for characterizing amorphous materials by X-ray diffraction

### 2018-04-17 doc/ms.tex
Escape _
Many edits for resubmission

\newcommand{\hypers}{\boldsymbol{\alpha}}
\newcommand{\params}{\boldsymbol{\theta}}
\newcommand{\data}{\boldsymbol{d}}
\newcommand{\info}{I}
\newcommand{\x}{x}
\newcommand{\todo}{\color{orange} \bf}

% http://support.river-valley.com/wiki/index.php?title=Elsarticle.cls

@@

\title{AMORPH: A statistical program for characterizing amorphous materials by X-ray diffraction}

@@

\begin{abstract}
AMORPH utilizes a new Bayesian statistical approach to interpreting X-ray diffraction results of samples with both crystalline and amorphous components. AMORPH fits X-ray diffraction patterns with a mixture of gaussians,
which allows us to simultaneously infer all of the model parameters and quantify their
uncertainties.
The program simulates background patterns previously manually applied, providing reproducible results, and significantly reducing inter- and intra-user biases. This approach allows for the quantification of amorphous and crystalline materials and for the characterization of the amorphous component, including parameters of centre of mass, width, skewness, and nongaussianity of the amorphous component. Results demonstrate the applicability of this program for calculating amorphous contents of volcanic materials and independently modeling their properties in compositionally variable materials.
\end{abstract}

\begin{keyword}
  Amorphous materials \sep X-ray diffraction \sep Bayesian inference \sep Markov Chain Monte Carlo \sep Mixture models \sep crystallinity
\end{keyword}

@@

\section{Introduction}
Quantification and characterization of amorphous materials is an important area of research in
terrestrial and planetary sciences \citep[e.g.,][]{schmidt2009, dehouck2014, wall2014, morris2016, zorn2018}.
Powder X-ray diffraction (XRD), although primarily a tool for characterization of crystalline materials,
also provides a means to investigate amorphous materials \citep[e.g.,][]{rowe2012, achilles2013}. X-ray amorphous materials lack long-range crystallographic order and produce broad humps with low intensities in XRD patterns. X-ray diffraction of rocks/soils containing amorphous materials (e.g. volcanic
glass, amorphous silica, and organics) produces a diffraction pattern which is a combination of the
crystalline and amorphous phases. Recent studies have debated the nature of amorphous materials in
planetary sciences \citep[e.g.][]{schmidt2009, dehouck2014}, and a growing number of
applications to terrestrial volcanics have presented an avenue of expansion for XRD methodologies
for amorphous material analysis \citep{ellis2015, kanakiya2017, zorn2018}.
Here we present a new statistical approach and computer program, AMORPH, which utilizes Bayesian inference to determine
the crystallinity, and other characteristics, of partially amorphous materials from X-ray diffraction patterns, as well as the uncertainty in these quantities.

Several techniques currently exist for the quantification of amorphous materials in heterogeneous (amorphous + crystalline) materials. However, these
each present their own limitations. The two most commonly used approaches in recent literature
either rely on a calibration curve for different crystallinities \citep[calibration method;][]{ rowe2012} or a combined Rietveld-Reference Intensity Ratio (RIR) method \citep{gualtieri1996, gualtieri2000}. While the combined Rietveld-RIR procedure allows for quantification of both the amorphous 
and crystalline materials, for proper characterization it requires a library of pure components,
spiked with a reference material (e.g. corundum). Sample spiking presents additional complications
as it reduces the observed intensity of the amorphous material thus reducing the accuracy of
pattern matching to determine abundances. An Excel based adaptation of this pattern matching
technique \citep[FULLPAT;][]{chipera2002} is routinely utilized for characterizing remotely acquired X-ray
diffraction patterns, such as those measured by the CheMin instrument on the Mars Science Laboratory (MSL) Curiosity rover \citep{blake2012, bish2013}. The problem faced by RIR approaches is they are based on modeling of an observed 
diffraction pattern compared to library measurements, and therefore do not allow an independent
assessment of the characteristics of the amorphous material. Furthermore, natural X-ray amorphous phases have a wide variety of compositions such that the library measurements may not include the amorphous phase(s) present in the sample. In comparison, the calibration method
focuses solely on the quantification of the amorphous material, utilizing the integrated counts
associated with the amorphous and crystalline componentry.

\begin{align}
C_{\rm m} \% = [C_{\rm pa}]/[A_{\rm pa} + C_{\rm pa}] \times 100
\end{align}

where $C_{\rm m}$ is the measured crystallinity, and $C_{\rm pa}$ and $A_{\rm pa}$ are the integrated peak areas for the
crystalline and amorphous components, respectively (see Figure 5 of \citet{wall2014}). 
While its simplicity has made it advantageous, the calibration methodology has been limited in that
the quality of the results largely depend on the ability of the user to manually assign
backgrounds for count integration, resulting in potentially large inter-user variations. Although the
point of the calibration curve is to reduce the inter-user variation, determinations still tend to
be associated with moderate user error (3-5 vol \%). As every user must independently practice
background fittings using the calibration method, the results ultimately are only as good as the
user’s ability to manually model the data. This technique, while not constrained by a
necessary library of diffraction patterns, also does not characterize the XRD properties of the
amorphous material. This study presents a new approach using Bayesian inference to automate
background determination for the calibration method, removing inter-user and intra-user biases in
the determination of crystallinity and peak characteristics for amorphous/partially amorphous
materials, regardless of what they are.

This paper is structured as follows. In Section~\ref{sec:bayes}, we outline the Bayesian
framework of inference in general terms, and in Section~\ref{sec:model} we describe the
specific model assumptions made by AMORPH (i.e., the mixture of gaussians), and
Section~\ref{sec:priors} lists the probabilistic assumptions (i.e., the prior distributions) of the program. Section~\ref{sec:applications} presents some
example applications and comparisons with standard techniques, and finally we conclude
in Section~\ref{sec:conclusions}. Basic instructions for installation and use of AMORPH are given in~\ref{sec:program}.

\section{Bayesian inference}\label{sec:bayes}
AMORPH is founded upon Bayesian inference
\citep{gregory2005bayesian, o2004kendall, sivia2006data}, where
probability distributions are used to model
states of uncertainty about the values of unknown quantities.
Bayesian inference is often implemented computationally with
Markov Chain Monte Carlo (MCMC) methods \citep{mackay2003}, and AMORPH is no exception.
Loosely speaking, the goal is to fit a dataset with a model by exploring
those regions of the model's parameter space that are consistent with the
data. The model assumptions are described in detail in
Section~\ref{sec:model}, and the computation is performed with
Diffusive Nested Sampling, an advanced MCMC method \citep{dns, dnest4}.
There have been many applications of Bayesian inference to
problems of separating background from compact signals
when noise is present \citep[e.g.][]{}. AMORPH can be viewed as another
case of this, with assumptions customized to the specific application.

Typically, one starts with the `prior distribution'
for unknown quantities (`parameters') $\params$, written $p(\params | \info)$
(the probability distribution for $\params$ {\em given} prior information
$\info$). Throughout this paper, $\params$ is used to denote a
collection of unknown parameters, whereas $2\theta$ (with the factor of two, and the
non-bold $\theta$) is later used to
refer to the incident angle plus reflected angle, which is the
standard notation in X-ray diffraction.

The prior describes an initial state of uncertainty about the
parameter values, and is often a wide probability distribution.
The prior is supplemented with $p(\data | \params, \info)$, sometimes
called a `sampling distribution',
which describes the uncertainty about the data which we {\em would} have,
if we knew the parameters (and the prior information), as a function of
the parameters. This describes uncertainty about what data will be observed,
but encodes some knowledge about the fact that the data have something to do with the
parameters $\params$. For example, $p(\data | \params, \info)$ often encodes
the assumption that the data has noise in it, and the specific noise values
are unknown a priori.

The
product of these two distributions is the joint prior
\begin{align}
p(\params, \data | \info) &= p(\params | \info)p(\data | \params, \info)
\end{align}
which describes uncertainty about the value of both the parameters and the
dataset. Once a specific observed dataset $\data_{\rm obs}$ is known, knowledge about
$\params$ is updated from the prior $p(\params | \info)$ to the posterior
\begin{align}
p(\params | \data_{\rm obs}, \info) &\propto
    p(\params | \info)p(\data_{\rm obs} | \params, \info).
\end{align}
which takes the specific data $\data_{\rm obs}$ into account. With the specific dataset,
$p(\data_{\rm obs} | \params,\info)$ becomes a function of the
unknown parameters $\params$ only and is called the {\em likelihood function}.
The posterior distribution is therefore proportional to the prior distribution
times the likelihood function.

When $\params$ is a large collection of parameters, the posterior distribution
is a probability distribution over a high-dimensional parameter space. To
represent it computationally, Markov Chain Monte Carlo (MCMC) methods are
often used to generate samples from the parameter space, according to the
posterior distribution. The output is a collection of plausible values of the
parameters which can be used to approximate posterior probabilities.
For example, the probability some parameter $\phi$ is greater than 3.4
is approximately the proportion of the posterior samples that have
$\phi > 3.4$. While many data analysis procedures work by finding the
best fitting values of the parameters (i.e., a maximum likelihood estimate),
Bayesian inference via MCMC involves finding a sampling of plausible values
for the parameters. Since AMORPH uses \citet{dnest4}, it also calculates
the `marginal likelihood' or `evidence' quantity $p(\data | \info)$, which
quantifies how well the model fits the data overall, averaged over all
possible values of its parameters. This allows straightforward model comparison
with any alternatives that may be proposed in the future
\citep{skilling2006nested}.

% BJB: Hopefully this addresses reviewer 1, point a.
The computational cost of running AMORPH on a data set of around 1000
points varies, as the posterior distribution depends on the data.
Another important factor is the number of MCMC samples desired.
On a modern PC, some useful posterior samples can be obtained in about
half an hour. With 300 posterior samples, typical analyses take approximately 7 hours to complete, depending on computing capabilities and the number of data points. However, the complete runs presented in this paper took
about a day each to run, as we wanted to ensure the results contained
about 1000 posterior samples.

@@

The model curve is a function $y = f(\x; \params)$,
which is assumed to be a sum of the following components identified in Figure~\ref{fig:example_data2}:
\begin{enumerate}
\item A background component, described by parameters $\params_{\rm bg}$
\item An amorphous component, described by parameters $\params_{\rm amorph}$
\item A crystalline component, described by parameters $\params_{\rm xtal}$
\end{enumerate}
The total model function $f_{\rm tot}(x)$ is therefore
\begin{align}
f_{\rm tot}(\x) &= f_{\rm bg}(\x; \params_{\rm bg})
       + f_{\rm amorph}(\x; \params_{\rm amorph})
       + f_{\rm xtal}(\x; \params_{\rm xtal})
\end{align}
and the measured amorphous content ($A_{\rm m}$ or $1-C_{\rm m}$) is defined, similar to Equation 1, as the proportion of the area in the
amorphous and crystalline components which is accounted for by the amorphous component:
\begin{align}
{A_{\rm m} = 1-C_{\rm m}} &= \frac{\int_{\x_{\rm min}}^{\x_{\rm max}}
            f_{\rm amorph}(\x; \params_{\rm amorph}) \, d\x}
          {\int_{\x_{\rm min}}^{\x_{\rm max}}
            f_{\rm amorph}(\x; \params_{\rm amorph}) \, dx
            + \int_{\x_{\rm min}}^{\x_{\rm max}} f_{\rm xtal}(\x; \params_{\rm xtal}) \, d\x} \label{eqn:crystallinity}
\end{align}

From each sample from the posterior distribution for all of the parameters,
it is possible to compute the proportion of amorphous material, making it straightforward to
calculate the marginal posterior distribution of the amorphous content, describing
the remaining uncertainty about its value in the light of the data.
In the following subsections, we describe the specific parameterizations of the
three parts of the model curve, that is, the functional forms of
$f_{\rm bg}$, $f_{\rm amorph}$, and $f_{\rm xtal}$.

\subsection{The background component}
The background component is assumed to be piecewise linear, with constant
slope
{between $\x=\x_{\rm min}$ and $\x=10$\degree, between $\x=10$\degree~and $\x=40$\degree,
and between $\x=40$\degree~and $\x=\x_{\rm max}$}. These values are selected to maintain a linear background over the interval from 10--40\degree, the predominant range of amorphous materials (8.8 -- 2.25 \AA) analyzed using a Cu K$\alpha$ x-ray source (Table~\ref{tab:anode}), while still allowing for some flexibility in the structure of the background outside this range.

We parameterized the background component using
five parameters. Firstly, there is a mean level $b$, which sets the
typical amplitude of the background component. There are also four parameters
$n_1, ..., n_4$, which determine the offset of four `control points'
from the mean level. The positions of the four control points are then:
\begin{align}
\left(\x_{\rm min}, be^{n_1}\right) \nonumber\\
\left(10, be^{n_2}\right) \nonumber\\
\left(40, be^{n_3}\right) \nonumber\\
\left(\x_{\rm max}, be^{n_4}\right). \nonumber
\end{align}
Control points may be modified, depending on the X-ray source in use and changes in the required range for the linear background as a function of the nature of the amorphous material (Table~\ref{tab:anode}). The value of the background component at any other point is found by
linear interpolation of these control points.
Figure~\ref{fig:background} shows three example
background curves generated from this parameterization, with typical
values of the parameters.

\begin{table}
\footnotesize
\centering
\begin{tabular}{|llll|}
\hline
{\bf X-ray Anode}      &   {\bf X-ray wavelength}   &  {\bf Control point 2} & {\bf Control point 3}\\
 &   {(\AA)}   &  {\bf(\degree $2\theta$)} & {\bf(\degree $2\theta$)}\\
\hline
Cr & 2.29 & 15.0 & 61.2\\
Fe & 1.94 & 12.7 & 51.1\\
Co & 1.79 & 11.7 & 46.9\\
{\bf Cu} & {\bf 1.54} & {\bf 10.0} & {\bf 40.0}\\
Mo & 0.71 & 4.6 & 18.2\\
Ag & 0.56 & 3.6 & 14.3\\
\hline
\end{tabular}
\caption{\it Typical XRD anodes with recommended background control points based on a predominant 8.8--2.25 {\AA}  amorphous peak.\label{tab:anode}}
\end{table}

@@

\subsection{The crystalline component}

% BJB: Reponse to Reviewer 2, comment about Lorentzian shape vs Gaussian
% We experimented with the Lorentzian shape instead of the Gaussian and found
% that for same datasets, it was preferred, whereas for other datasets, the
% Gaussian is preferred (as adjudicated by the marginal likelihood values
% for the two different assumptions). Rather than having to run each model
% twice, we decided it would be more elegant to replace the gaussian assumption
% with some family of shapes that can smoothly interpolate between gaussian
% and lorentzian as necessary. Specifically, we chose the Student's
% t-distribution which can do this straightforwardly. This introduces one
% more parameter to the model, which controls the shape of the peaks.

% I (BJB) also changed the parameterisation of the heights of the peaks, so
% they are described by the total integrated mass rather than the peak
% amplitude. This makes the Metropolis exploration a bit more efficient.

We assume that there is some number $N_{\rm xtal}$ of narrow peaks
due to crystalline material,
each of which has the same shape. The shapes of the narrow
peaks are controlled by a single shape parameter $\nu_{\rm xtal}$,
assumed common to all of the narrow peaks. The chosen shape is based
on the Student's $t$-distribution, which allows to peaks to be heavy-tailed
if $\nu_{\rm xtal}$ is low, and to be Lorentzian or Cauchy in shape if
$\nu_{\rm xtal}$ is exactly 1. As $\nu_{\rm xtal}$ is high, the shape
approximates a gaussian.

The $i$th peak is parameterized
by its central position $X_{i, \rm xtal}$,
total integrated `mass' $M_{i, \rm xtal}$, and width $W_{i, \rm xtal}$.
The parameter vector for crystalline component is
\begin{align}
\params_{\rm xtal} &=
  \{N_{\rm xtal}, \nu_{\rm xtal}, X_{1, \rm xtal}, ..., X_{N, \rm xtal},
    M_{1, \rm xtal}, ..., M_{N, \rm xtal},
    W_{1, \rm xtal}, ..., W_{N, \rm xtal}\}
\end{align}
and the total shape contributed by the
crystalline component is a sum of the $N_{\rm xtal}$ peaks:
\begin{align}
f_{\rm xtal}(\x; \params_{\rm xtal}) &=
	\frac{\Gamma(\frac{\nu_{\rm xtal} + 1}{2})}{\sqrt{\nu_{\rm xtal}\pi}
                 		\Gamma(\frac{\nu_{\rm xtal}}{2})}
    \sum_{i=1}^{N_{\rm xtal}}
    		\frac{M_{i, \rm xtal}}
            	 {W_{i, \rm xtal}}
	\left(1 + \frac{1}{\nu}\left(\frac{x - X_{i, \rm xtal}}{W_{i, \rm xtal}}\right)^2\right)
    			^{-\frac{\nu_{\rm xtal} + 1}{2}}. \label{eqn:t_shape}
\end{align}
The terms inside the sum have the form of the probability density function
of the $t$-distribution, adjusted by shifting according to the central
position and widening by the width, and the total integral is scaled by
the mass parameters. $\Gamma()$ is the gamma function.

For computational efficiency, evaluation of Equation~\ref{eqn:t_shape}
is performed using a lookup table.

\subsection{The amorphous component}
The amorphous component is also composed of a sum of peaks, which we chose to be
gaussians. The parameterization is the same as for the crystalline peaks but with a fixed,
high value for the shape parameter.
Apart from the gaussian shape, the only other difference is the prior distributions for the positions, masses, and widths, given in Section~\ref{sec:priors}. In short, the wide gaussians representing the
amorphous component are more likely to be located near the middle of the
dataset, and to be wider.

>>>>

\newcommand{\hypers}{\boldsymbol{\alpha}}
\newcommand{\params}{\boldsymbol{\theta}}
\newcommand{\data}{\boldsymbol{d}}
\newcommand{\info}{I}
\newcommand{\x}{x}
\newcommand{\todo}{\color{orange} \bf}
\newcommand{\mnras}{Monthly Notices of the Royal Astronomical Society}
\newcommand{\changed}{\color{blue}}
\renewcommand{\thefootnote}{\arabic{footnote}}
% http://support.river-valley.com/wiki/index.php?title=Elsarticle.cls

@@

\title{AMORPH: A statistical program for characterizing amorphous materials by X-ray diffraction$^1$\footnotetext{\changed $^1$MR conceived the project, developed geologic applications, and drafted the manuscript with BB. BB developed the statistical model, wrote the program code, and co-wrote the manuscript.}}

@@

\begin{abstract}
AMORPH utilizes a new Bayesian statistical approach to interpreting X-ray diffraction results of {\changed samples with both crystalline and amorphous components}. AMORPH fits X-ray diffraction patterns with a mixture of {\changed narrow and wide components},
{\changed simultaneously inferring} all of the model parameters and {\changed quantifying} their
uncertainties.
The program simulates background patterns previously {\changed applied manually}, providing reproducible results, and significantly reducing inter- and intra-user biases. This approach allows for the quantification of amorphous and crystalline materials and for the characterization of the amorphous component, including {\changed properties such as the} centre of mass, width, skewness, and nongaussianity of the amorphous component. Results demonstrate the applicability of this program for calculating amorphous contents of volcanic materials and independently modeling their properties in compositionally variable materials.
\end{abstract}

\begin{keyword}
  Amorphous materials \sep X-ray diffraction \sep Bayesian inference \sep Markov chain monte carlo \sep mixture models \sep crystallinity
\end{keyword}

@@

\section{Introduction}
Quantification and characterization of amorphous materials is an important area of research in
terrestrial and planetary sciences \citep[e.g.,][]{schmidt2009, dehouck2014, wall2014, morris2016, zorn2018}.
Powder X-ray diffraction (XRD), although primarily a tool for characterization of crystalline materials,
also provides a means to investigate amorphous materials \citep[e.g.,][]{rowe2012, achilles2013}. X-ray {\changed amorphous materials lack long-range crystallographic order and produce broad humps with low intensities in XRD patterns. X-ray} diffraction of rocks/soils containing amorphous materials (e.g. volcanic
glass, amorphous silica, and organics) produces a diffraction pattern which is a combination of the
crystalline and amorphous phases. Recent studies have debated the nature of amorphous materials in
planetary sciences \citep[e.g.][]{schmidt2009, dehouck2014}, and a growing number of
applications to terrestrial volcanics have presented an avenue of expansion for XRD methodologies
for amorphous material analysis \citep{ellis2015, kanakiya2017, zorn2018}.
Here we present a new statistical approach and computer program, AMORPH, which utilizes Bayesian inference to determine
the crystallinity, and other characteristics, of partially amorphous materials from X-ray diffraction patterns, as well as the uncertainty in these quantities.

Several techniques currently exist for the quantification of amorphous materials in heterogeneous (amorphous + crystalline) materials. However, these
each present their own limitations. The two most commonly used approaches in recent literature
either rely on a calibration curve for different crystallinities \citep[calibration method;][]{ rowe2012} or a combined Rietveld-Reference Intensity Ratio (RIR) method \citep{gualtieri1996, gualtieri2000}. While the combined Rietveld-RIR procedure allows for quantification of both the amorphous 
and crystalline materials, for proper characterization it requires a library of pure components,
spiked with a reference material (e.g. corundum). Sample spiking presents additional complications
as it reduces the observed intensity of the amorphous material thus reducing the accuracy of
pattern matching to determine abundances. An Excel-based adaptation of this pattern matching
technique \citep[FULLPAT;][]{chipera2002} is routinely utilized for characterizing remotely acquired X-ray
diffraction patterns, such as those {\changed measured by the CheMin instrument on} the Mars Science Laboratory (MSL) Curiosity rover \citep{blake2012, bish2013}. The problem faced by RIR approaches is they are based on modeling of an observed 
diffraction pattern compared to library measurements, and therefore do not allow an independent
assessment of the characteristics of the amorphous material. {\changed Furthermore, natural X-ray amorphous phases have a wide variety of compositions such that the library measurements may not include the amorphous phase(s) present in the sample.} In comparison, the calibration method
focuses solely on the quantification of the amorphous material. {\changed This method utilizes} the integrated counts
associated with the amorphous and crystalline componentry:

\begin{align}
C_{\rm m} \% = [C_{\rm pa}]/[A_{\rm pa} + C_{\rm pa}] \times 100\label{eqn:definition}
\end{align}

where $C_{\rm m}$ is the measured crystallinity, and $C_{\rm pa}$ and $A_{\rm pa}$ are the integrated peak areas for the
crystalline and amorphous components, respectively (see Figure 5 of \citet{wall2014}). 
While its simplicity has made it advantageous, the calibration methodology has been limited in that
the quality of the results largely depend on the ability of the user to manually assign
backgrounds for count integration, resulting in potentially large inter-user variations. Although the
point of the calibration curve is to reduce the inter-user variation, determinations still tend to
be associated with moderate user error (3--5 vol \%). As every user must independently practice
background fittings using the calibration method, the results ultimately are only as good as the
user’s ability to manually {\changed fit} the data. This technique, while not constrained by a
necessary library of diffraction patterns, also does not characterize the XRD properties of the
amorphous material. This study presents a new approach using Bayesian inference to automate
background determination for the calibration method, removing inter-user and intra-user biases in
the determination of crystallinity and peak characteristics for amorphous/partially amorphous
materials, regardless of what they are.

This paper is structured as follows. In Section~\ref{sec:bayes}, we outline the Bayesian
framework of inference in general terms. In Section~\ref{sec:model} we describe the
specific modelling assumptions made by AMORPH, and
Section~\ref{sec:priors} lists the probabilistic assumptions (i.e., the prior distributions)
{\changed assumed by} the program. Section~\ref{sec:applications} presents some
example applications and comparisons with standard techniques, and finally we conclude
in Section~\ref{sec:conclusions}. Basic instructions for installation and use of AMORPH are given in~\ref{sec:program}.

\section{Bayesian inference}\label{sec:bayes}
AMORPH is founded upon Bayesian inference
\citep{gregory2005bayesian, o2004kendall, sivia2006data}, where
probability distributions are used to model
states of uncertainty about the values of unknown quantities.
Bayesian inference is often implemented computationally with
Markov Chain Monte Carlo (MCMC) methods \citep{mackay2003}, and AMORPH is no exception.
Loosely speaking, the goal is to fit a dataset with a model by exploring
those regions of the model's parameter space that are consistent with the
data. The model assumptions are described in detail in
Section~\ref{sec:model}, and the computation is performed with
Diffusive Nested Sampling {\changed \citep[an advanced MCMC method][]{dns}
as implemented in the software package DNest4 \citep{dnest4}.
This method is robust to complicated posterior distributions.

There have been many approaches to
problems of separating background from compact signals
when noise is present, such as the famous CLEAN algorithm \citep{clean},
through to more explicitly Bayesian approaches
\citep[e.g.][]{padayachee, massinf, object, hobson, worpel}.
AMORPH can be viewed as another
case of this, with assumptions customized to this specific application.}

Typically, one starts with the `prior distribution'
for unknown quantities (`parameters') $\params$, written $p(\params | \info)$
(the probability distribution for $\params$ {\em given} prior information
$\info$). Throughout this paper, the bold $\params$ is used to denote a
collection of unknown parameters, whereas $2\theta$ (with the factor of two, and the
non-bold $\theta$) is later used to
refer to the incident angle plus reflected angle, which is the
standard notation in X-ray diffraction.

The prior describes an initial state of uncertainty about the
parameter values, and is often a wide probability distribution.
The prior is supplemented with $p(\data | \params, \info)$, sometimes
called a `sampling distribution',
which describes the uncertainty about the data which we {\em would} have,
if we knew the parameters (and the prior information), as a function of
the parameters. {\changed The sampling distribution} describes uncertainty about what data will be observed,
but encodes some knowledge about the fact that the data have something to do with the
parameters $\params$. For example, $p(\data | \params, \info)$ often encodes
the assumption that the data has noise in it, and the specific noise values
are unknown a priori.

The
product of these two distributions is the joint prior
\begin{align}
p(\params, \data | \info) &= p(\params | \info)p(\data | \params, \info)
\end{align}
which describes uncertainty about the value of both the parameters and the
dataset. Once a specific observed dataset $\data_{\rm obs}$ is known, knowledge about
$\params$ is updated from the prior $p(\params | \info)$ to the posterior
\begin{align}
p(\params | \data_{\rm obs}, \info) &\propto
    p(\params | \info)p(\data_{\rm obs} | \params, \info).
\end{align}
which takes the specific data $\data_{\rm obs}$ into account {\changed via conditioning}.
With the specific dataset,
$p(\data_{\rm obs} | \params,\info)$ becomes a function of the
unknown parameters $\params$ only and is called the {\em likelihood function}.
The posterior distribution is therefore proportional to the prior distribution
times the likelihood function.

When $\params$ is a large collection of parameters, the posterior distribution
is a probability distribution over a high-dimensional parameter space. To
represent it computationally, Markov Chain Monte Carlo (MCMC) methods are
often used to generate samples from the parameter space, according to the
posterior distribution. The output is a collection of plausible values of the
parameters which can be used to approximate posterior probabilities.
For example, the probability some parameter $\phi$ is greater than 3.4
is approximately the proportion of the posterior samples that have
$\phi > 3.4$. While many data analysis procedures work by finding the
best fitting values of the parameters (i.e., a maximum likelihood estimate),
Bayesian inference via MCMC involves {\changed exploring a range of plausible values
for the parameters. Since AMORPH uses Diffusive Nested Sampling, it also calculates
the `marginal likelihood' or `evidence' quantity $p(\data | \info)$, which
quantifies how well the model fits the data overall, averaged over all
possible values of its parameters. This allows straightforward model comparison
with any alternatives that may be proposed in the future
\citep{skilling2006nested}.

The computational cost of running AMORPH
varies, as the posterior distribution depends on the data.
The size of the dataset is important, of course.
Another important factor is the number of MCMC samples desired.
To quantify the uncertainty about a few scalar parameters, about
100 posterior samples is usually sufficient. Indeed, \citet[][p. 380]{mackay2003} suggests
that only about twelve are needed for the uncertainty about a single
quantity!
However, if one wants to produce
smooth histograms or look at details of the posterior distribution,
many more samples are required.

On a modern desktop PC and using aggressive numerical settings
(see~\ref{sec:program} for details),
some useful posterior samples can be obtained in about
15 minutes when the dataset contains about 1500 points.
Typical analyses take approximately 5 hours to fully complete and yield
a few hundred posterior samples.
However, the complete runs presented in this paper took
about a day each to run, as we ran with more conservative settings
to ensure the number of posterior samples was about 1000 for each dataset.}

@@

The model curve is a function $y = f(\x; \params)$,
which is assumed to be a sum of the following components identified in Figure~\ref{fig:example_data2}:
\begin{enumerate}
\item A background component, described by parameters $\params_{\rm bg}$
\item An amorphous component, described by parameters $\params_{\rm amorph}$
\item A crystalline component, described by parameters $\params_{\rm xtal}$
\end{enumerate}
The total model function $f_{\rm tot}(x)$ is therefore
\begin{align}
f_{\rm tot}(\x) &= f_{\rm bg}(\x; \params_{\rm bg})
       + f_{\rm amorph}(\x; \params_{\rm amorph})
       + f_{\rm xtal}(\x; \params_{\rm xtal})
\end{align}
and the measured amorphous content ($A_{\rm m}$ or $1-C_{\rm m}$) is defined,
similar to Equation~\ref{eqn:definition}, as the proportion of the area in the
amorphous and crystalline components which is accounted for by the amorphous component:
\begin{align}
{A_{\rm m} = 1-C_{\rm m}} &= \frac{\int_{\x_{\rm min}}^{\x_{\rm max}}
            f_{\rm amorph}(\x; \params_{\rm amorph}) \, d\x}
          {\int_{\x_{\rm min}}^{\x_{\rm max}}
            f_{\rm amorph}(\x; \params_{\rm amorph}) \, dx
            + \int_{\x_{\rm min}}^{\x_{\rm max}} f_{\rm xtal}(\x; \params_{\rm xtal}) \, d\x} \label{eqn:crystallinity}
\end{align}

From each sample from the posterior distribution for all of the parameters,
it is straightforward to compute the proportion of amorphous material, and hence
calculate the marginal posterior distribution of the amorphous content.
In the following subsections, we describe the specific parameterizations of the
three parts of the model curve, that is, the functional forms of
$f_{\rm bg}$, $f_{\rm amorph}$, and $f_{\rm xtal}$.

\subsection{The background component}
The background component is assumed to be piecewise linear, with constant
slope
{between $\x=\x_{\rm min}$ and $\x=10$\degree, between $\x=10$\degree~and $\x=40$\degree,
and between $\x=40$\degree~and $\x=\x_{\rm max}$}. These values are selected to maintain a linear background over the interval from 10--40\degree, the predominant range of amorphous materials {\changed (8.8 -- 2.25 \AA)} analyzed using a Cu K$\alpha$ x-ray source {\changed (Table~\ref{tab:anode})}, while still allowing for some flexibility in the structure of the background outside this range.

We parameterized the background component using
five parameters. Firstly, there is a mean level $b$, which sets the
typical amplitude of the background component. There are also four parameters
$n_1, ..., n_4$, which determine the offset of four `control points'
from the mean level. The positions of the four control points are then:
\begin{align}
\left(\x_{\rm min}, be^{n_1}\right) \nonumber\\
\left(10, be^{n_2}\right) \nonumber\\
\left(40, be^{n_3}\right) \nonumber\\
\left(\x_{\rm max}, be^{n_4}\right). \nonumber
\end{align}
{\changed Control points may be modified by editing the file {\tt control\_points.in},
depending on the X-ray source in use and changes in the required range for the linear background as a function of the nature of the amorphous material (Table~\ref{tab:anode}).} The value of the background component at any other point is found by
linear interpolation of these control points.
Figure~\ref{fig:background} shows three example
background curves generated from this parameterization, with typical
values of the parameters.

\begin{table}
\footnotesize
\centering
{\changed
\begin{tabular}{|llll|}
\hline
{\bf X-ray Anode}      &   {\bf X-ray wavelength}   &  {\bf Control point 2} & {\bf Control point 3}\\
 &   {(\AA)}   &  {\bf(\degree $2\theta$)} & {\bf(\degree $2\theta$)}\\
\hline
Cr & 2.29 & 15.0 & 61.2\\
Fe & 1.94 & 12.7 & 51.1\\
Co & 1.79 & 11.7 & 46.9\\
{\bf Cu} & {\bf 1.54} & {\bf 10.0} & {\bf 40.0}\\
Mo & 0.71 & 4.6 & 18.2\\
Ag & 0.56 & 3.6 & 14.3\\
\hline
\end{tabular}
\caption{\it Typical XRD anodes with recommended background control points based on a predominant 8.8--2.25 {\AA}  amorphous peak.\label{tab:anode}}
} % \changed
\end{table}

@@

\subsection{The crystalline component}
We assume that there is some number $N_{\rm xtal}$ of narrow peaks
{\changed due to crystalline material},
each of which has {\changed the same shape. The shapes of the narrow
peaks are controlled by a single shape parameter $\nu_{\rm xtal}$,
assumed common to all of the narrow peaks. The chosen shape is based
on the Student's $t$-distribution, which allows the peaks to be heavy-tailed
if $\nu_{\rm xtal}$ is low, and to be Lorentzian or Cauchy in shape if
$\nu_{\rm xtal}$ is exactly 1. As $\nu_{\rm xtal}$ is high, the shape
approximates a gaussian.}

The $i$th peak is parameterized
by its central position $X_{i, \rm xtal}$,
{\changed total integrated `mass' $M_{i, \rm xtal}$}, and width $W_{i, \rm xtal}$.
The parameter vector for crystalline component is {\changed
\begin{align}
\params_{\rm xtal} &=
  \{N_{\rm xtal}, \nu_{\rm xtal}, X_{1, \rm xtal}, ..., X_{N, \rm xtal},
    M_{1, \rm xtal}, ..., M_{N, \rm xtal},
    W_{1, \rm xtal}, ..., W_{N, \rm xtal}\}
\end{align} } % changed
and the total shape contributed by the
crystalline component is a sum of the $N_{\rm xtal}$ {\changed peaks:
\begin{align}
f_{\rm xtal}(\x; \params_{\rm xtal}) &=
	\frac{\Gamma(\frac{\nu_{\rm xtal} + 1}{2})}{\sqrt{\nu_{\rm xtal}\pi}
                 		\Gamma(\frac{\nu_{\rm xtal}}{2})}
    \sum_{i=1}^{N_{\rm xtal}}
    		\frac{M_{i, \rm xtal}}
            	 {W_{i, \rm xtal}}
	\left(1 + \frac{1}{\nu}\left(\frac{x - X_{i, \rm xtal}}{W_{i, \rm xtal}}\right)^2\right)
    			^{-\frac{\nu_{\rm xtal} + 1}{2}}. \label{eqn:t_shape}
\end{align}
The terms inside the sum have the form of the probability density function
of the $t$-distribution, adjusted by shifting according to the central
position and widening by the width, and the total integral is scaled by
the mass parameters. $\Gamma()$ is the gamma function.

For computational efficiency, evaluation of Equation~\ref{eqn:t_shape}
is performed using a lookup table. In addition, we evaluate the peak
shapes at the $x$-position of each data point, rather than integrating
the peak shape across a finite bin width. This is faster but must be
kept in mind when interpreting parameters affected by this assumption,
such as the widths of the narrow peaks. } %changed

\subsection{The amorphous component}
The amorphous component is also composed of {\changed a sum of peaks, which we chose to be
gaussians. The parameterization is the same as for the crystalline peaks but with a fixed,
high value for the shape parameter.
Apart from the gaussian shape, the only other difference is the prior distributions for the positions, masses, and widths, given in Section~\ref{sec:priors}. In short, the wide gaussians representing the
amorphous component are more likely to be located near the middle of the
dataset, and to be wider.
The constrainted prior for the positions of the wide amorphous
peaks prevents the program from attempting to explain background
variations near the edge of the data using the amorphous component.}
